---
title: "R Tutorial"
author: "Derek Lukacsko"
date: "September 8, 2016"
output: html_document
---

This part will be provided
```{r Setting Up Environment}
# Directory
set_wd(file.path = "")

# Load packages
packages <- c("tidyr","dplyr","ggplot2")
sapply(packages, library, character.only = TRUE)

# Working dataframe
df <- iris

##ADD AN NA AND CHANGE VARIABLE TYPES##
df$Sepal.Length <- as.character(df$Sepal.Length)
df$Species <- as.character(df$Species)
df$Sepal.Width[100] <- NA


# Inspect the data
dim(df)
colnames(df)
View(head(df))

```

1. Identify the feature types
2. Ensure all continous variables are type:numeric and categorical variables are type:factor
3. Identify where there are missing values
4. Handle the missing values
```{r Exercise 1: Data Cleaning}
# Feature/vairable types
sapply(df,class)

# Different ways to do the same thing
df[,"Sepal.Length"] <- as.factor(df$Sepal.Length)
df[,5] <- as.factor(df$Species)
df <- df %>%
  mutate(Sepal.Length = as.factor(Sepal.Length),
         Species = as.factor(Species))


# Number of missing values
sum(complete.cases(df))

# What column is it in
sapply(df, function(x) any(is.na(x)))

# What row is it in
df[is.na(df$Sepal.Width),]

# Remove the row manually

# Remove the row the easy way

# Replace the row as the median or mean 

```


1. Plot the relationship between variables Sepal.Length and Sepal.Width
2. Color the points by Species
3. Add a title
```{r Exercise 2: Data Inspection}

# Dot plot with linear fit
ggplot(df, aes(x=Sepal.Length,y=Sepal.Width)) + 
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship Between Sepal.Length and Sepal.Width")

# Dot plot colored by species with linear fit
ggplot(df, aes(x=Sepal.Length,y=Sepal.Width,color = Species)) + 
  geom_jitter() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship Between Sepal.Length and Sepal.Width \n by Species")

# Dot plot colored by species with linear fit, faceted by species
ggplot(df, aes(x=Sepal.Length,y=Sepal.Width,color = Species)) + 
  facet_wrap(~Species, nrow = 2) + #change nrow
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship Between Sepal.Length and Sepal.Width \n Faceted by Species")

# Same as previous but with all point present in each plot
ggplot(df, aes(Sepal.Length,Sepal.Width, color = Species)) +
  geom_point(data = transform(df, Species = NULL), colour = "grey85") +
  geom_point() +
  geom_smooth(method = "lm", se =FALSE) +
  facet_wrap(~Species, nrow= 2) +
 labs(title = "Relationship Between Sepal.Length and Sepal.Width \n Faceted by Species")

```

1. Make a box plot of Sepal.Length
2. Identify the rows that contain outliers
3. Handle the outliers
```{r Exercise 3: Identify Outliers}

```


1. Determine if Sepal.Length and Sepal.Width follow a Gaussian distribution using a test of normality
2. Make a histogram or frequency density plot to roughly visalize the feature distributions
3. Determine if the correlation between Sepal.Length and Sepal.Width is significant
```{r Exercise 4: test normality, test correlation}

ggplot(df2, aes(Sepal.Width)) + 
  stat_bin(bins = 20,  fill="steelblue", color = "white") + 
  theme_bw()

ggplot(df2, aes(Sepal.Width)) + 
  geom_density(fill= "steelblue") + 
  theme_bw()

colnames(df2)
df2 <- df[,c(1:4)]
cor.test(df2)

# Low P --> reject H0
cor.test(df2$Sepal.Length,df2$Sepal.Width, method = c("pearson"), conf.level = 0.95)[3]


```


